
# Training Materials on Cross-Validation

Cross-validation is not one-size-fits-all it depends on the structure of your data

## Why cross-validation matters

Goal: estimate generalization performance

Problem: single train/test split → high variance

Solution: resampling strategies

Key message: The way you split your data can make your model look better or worse than it really is.


```{r training setup}
library(rsample)
library(dplyr)
library(arrow)
```

```{r read file cv0}
data_cv0 <-  read_parquet("data/data_synth_cv0.parquet", as_data_frame = TRUE)

str(data_cv0)
```
File description:
This is a syntehtic file that simulates the answers received to the Survey on Construction Enterprises to collect key information on the structure and performance of the construction sector. Survey responses are used to define stratification variables that underpin estimation procedures extending beyond the sampled units. The effectiveness of this methodology critically depends on the completeness of all variables used in the stratum definition. However, non-response and partial non-response persist for a subset of these variables, posing challenges to data quality, accuracy, and reliability thus the use case in WP9 tries to predict the variables required for the strata.
- ANO is year,
- the other variables from DD_CC_FF_COD to UNI characterize the enterprise 
- IMP 0 for a real answer, 1 or 2 imputed with values from year n-1 or -2, 3 other imputations - the 3 variables started with I are Integrated administrative variables - the survey questions are prefixes with QF - source informs about the survey - ID is the anonymous ID of the enterprise - unit is the anonymous unit of the enterprise when it may have several units - the target is QF185001


