
# Training Materials on Cross-Validation

Cross-validation is not a one-size-fits-all technique — it should always be adapted to the structure and nature of your data.

---

## Why Cross-Validation Matters

The main goal of cross-validation is to obtain a reliable estimate of a model’s **generalization performance**, i.e., how well it will perform on unseen data.

A single train/test split is often not enough because it can lead to **high variance** in performance estimates. Depending on how the data is split, the model may appear artificially strong or weak.

To address this issue, we use **resampling strategies**, which repeatedly split the data in different ways and average the results to get a more stable evaluation.

---

## Key Insight

The way you split your data directly influences the perceived performance of your model.

A poorly chosen validation strategy can make a model look better (or worse) than it truly is in practice, leading to misleading conclusions.


```{r training setup}
library(rsample)
library(dplyr)
library(arrow)
```

```{r read file cv0}
data_cv0 <-  read_parquet("data/data_synth_cv0.parquet", as_data_frame = TRUE)

str(data_cv0)
```
# Construction Enterprises Survey Dataset (Synthetic File Description)

## Overview

This dataset is a synthetic file designed to simulate responses from the *Survey on Construction Enterprises*. The survey collects key information about the structure and performance of enterprises within the construction sector.

The resulting data are used to define **stratification variables**, which form the basis for estimation procedures that extend beyond the sampled units. The reliability of these estimation methods depends critically on the completeness and quality of all variables involved in the stratification process.

However, the dataset reflects real-world challenges such as **non-response** and **partial non-response**, which affect a subset of key variables. These issues introduce challenges related to data quality, accuracy, and overall reliability. To address this, the use case in WP9 focuses on predicting the variables required for stratification.

## Key Variables

### Time and Identifiers
- ANO: Reference year of the observation  
- ID: Anonymous identifier of the enterprise  
- UNIT: Anonymous identifier of the enterprise unit (used when an enterprise has multiple units)

### Enterprise Characteristics
Variables ranging from DD_CC_FF_COD to UNI describe structural and operational characteristics of each enterprise.

### Imputation Indicator
- IMP: Indicates the origin of the data value:
  - 0 → Real (observed) response  
  - 1 or 2 → Imputed values based on year n-1 or n-2  
  - 3 → Other types of imputation  

### Integrated Administrative Variables
- Variables prefixed with I* represent integrated administrative data sources.

### Survey Variables
- Variables prefixed with QF* correspond to survey questions.

## Target Variable

- QF185001: The target variable of interest in the WP9 use case

## Notes

This dataset is specifically designed to support methodological development in handling missing data and improving prediction of stratification variables. It reflects both observed and imputed data conditions commonly encountered in large-scale enterprise surveys.

