
## Cross-validation based on data structure

### Imbalanced data

Goal: Ensure reliable evaluation of model performance when predicting an imbalanced target

Problem: In imbalanced datasets, random folds may not contain enough observations of rare target values, leading to biased or misleading performance estimates

Solution: Stratified K-Fold Cross-Validation, which preserves the distribution of the target variable (QF185001) in each fold

Key message: If you don't preserve the target distribution, you are no longer evaluating the same problem in each fold

```{r Stratified K-Fold}
library(rsample)

set.seed(123)

folds <- vfold_cv(data_cv0, v = 5, strata = QF185001)

split1 <- folds$splits[[1]]

train_data <- analysis(split1)
test_data  <- assessment(split1)
```

in the case where the variable is continuous like in our case we can use bins

```{r Stratified K-Fold with Bins}

# create the binned varaiable
data_cv0 <- data_cv0 |>
  mutate(QF185001_bin = ntile(QF185001, 5))

# lighter inspection
dplyr::glimpse(data_cv0, width = 60)

folds <- vfold_cv(data_cv0, v = 5, strata = QF185001_bin)
```

```{r examine bins}
data_cv0 %>%
  group_by(QF185001_bin) %>%
  summarise(
    min = min(QF185001, na.rm = TRUE),
    max = max(QF185001, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(QF185001_bin)
```

```{r visualize distribution}
library(ggplot2)

ggplot(data_cv0, aes(x = factor(QF185001_bin), y = QF185001 + 1)) +
  geom_boxplot() +
  scale_y_log10()
```

### Small datasets

Goal: Ensure reliable estimation of model performance when data is limited

Problem: With small datasets, model evaluation has high variance because each split contains very few observations, making results unstable and sensitive to data partitioning

Solution: - Leave-One-Out Cross-Validation (LOOCV): maximizes training data usage - Repeated K-Fold Cross-Validation: reduces variance by averaging across multiple random splits

Key Message: When data is scarce, the main risk is not bias but high variance in evaluation. Use resampling strategies that stabilize results

#### Leave-One-Out Cross-Validation (LOOCV)

uses all data efficiently but does not scale well, as it requires training one model per observation.

```{r LOOCV Step 1 - What are the folds?}

library(rsample)

# keeping only numeric predictors in the example 
data_num <- data_cv0 %>% select(where(is.numeric))

# making a small sample because we will have as many
# folds as observations
data_small <- data_num %>% slice_sample(n = 100) 

folds <- loo_cv(data_small)

length(folds$splits)

```

```{r LOOCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

nrow(train)
nrow(test)
```

```{r LOOCV Step 3 - How do we compute error?}

# errs stores the error for each fold
# since each fold has 1 test observation, each value is
# the prediction error for that observation
errs <- sapply(folds$splits, function(s) {
  train <- analysis(s)
  test  <- assessment(s)
  
  # lm is simple example model (linear regression)
  # not required for LOOCV
  # may replace it with any model (tree, random forest, etc.)
  m <- lm(QF185001 ~ ., data = train)
  pred <- predict(m, test)
  
  # mean squared error
  mean((test$QF185001 - pred)^2)
})
```

```{r}
# train many models (almost identical)
# each tested on one observation
# then average the errors
mean(errs)
```

#### Repeated K-Fold Cross-Validation

Repeated K-Fold reduces variance by averaging over multiple data splits.

```{r RKFCV Step 1 - What are the folds?}

library(rsample)

# split data into 5 folds
# repeat the whole process 10 times with different random splits
# Total models trained = 5 × 10 = 50
folds <- vfold_cv(data_cv0, v = 5, repeats = 10)

nrow(folds)

folds$splits[[1]]
```

```{r RKFCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

nrow(train)
nrow(test)
```

Train/test separation (\~80% / 20%)

```{r RKFCV Step 3 - How do we compute error?}

library(rpart)

# train many models on different splits
# compute error each time
# and average provide more stable estimate
errs <- sapply(folds$splits, function(s) {
  train <- analysis(s)
  test  <- assessment(s)

  train$ESTRATO <- factor(train$ESTRATO)
test$ESTRATO  <- factor(test$ESTRATO, levels = levels(train$ESTRATO))
  train$DD_CC_FF_COD <- factor(train$DD_CC_FF_COD)
  test$DD_CC_FF_COD  <- factor(test$DD_CC_FF_COD,
                               levels = levels(train$DD_CC_FF_COD))

  m <- rpart(QF185001 ~ ., data = train)
  pred <- predict(m, test)

  mean((test$QF185001 - pred)^2)
})

mean(errs)

```

```{r read file cv1}
data_cv1 <- read_parquet("data/data_synth_cv1.parquet")

dplyr::glimpse(data_cv1)
```

### Grouped / clustered data

Goal: Ensure that model evaluation does not leak information across related observations

Problem: Observations from the same entity (e.g. entreprise) are not independent If split randomly, the same entreprise can appear in both train and test, i.e. data leakage

Solution: Use Group K-Fold Cross-Validation, ensuring that all observations from the same group stay in the same fold

Key Message: If groups are split across folds, the model is evaluated on data it has effectively already seen

Examples: - multiple rows per user - entreprises, households, devices, patients

#### Group K-Fold Cross-Validation

Same entreprise appears multiple times, requires clustered data - ID - entreprise (group) - unit - establishment (multiple per entreprise)

```{r GKFCV Step 1 - What are the folds?}
library(rsample)

folds <- group_vfold_cv(data_cv1, group = ID, v = 5)
```

```{r GKFCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

head(unique(train$ID), 10)
head(unique(test$ID), 10)

# check no overlap in groups
intersect(unique(train$ID), unique(test$ID))
```

For contrast we will compare with Random CV where we will see overlap

```{r Random CV}
folds_random <- vfold_cv(data_cv1, v = 5)

s <- folds_random$splits[[1]]

train_r <- analysis(s)
test_r  <- assessment(s)

intersect(unique(train_r$ID), unique(test_r$ID))[1:50]
```
  
### Hierarchical / multi-level data

Examples: establishments within entreprises, individuals within households, students within schools Solutions: Nested grouping strategies combine: Group CV + Stratification

Problem: rsample does not natively support group + strata simultaneously How: 1 - Stratify at group level 2 - Split at group level 3 - Apply at observation level

```{r stratify at enterprise level}
library(dplyr)
data_synth<-data_cv1
df_grouped <- data_synth %>%
  group_by(ID) %>%
  summarise(
    QF185001_mean = mean(QF185001, na.rm = TRUE)
  ) %>%
  mutate(strata = ntile(QF185001_mean, 5))

dplyr::glimpse(df_grouped, width = 60)
```

```{r Split at enterprise level using strata}
library(rsample)

folds_groups <- vfold_cv(df_grouped, v = 5, strata = strata)
```

```{r Apply at unit level}
folds <- lapply(folds_groups$splits, function(s) {
  
  train_ids <- analysis(s)$ID
  test_ids  <- assessment(s)$ID
  
  list(
    train = data_synth %>% filter(ID %in% train_ids),
    test  = data_synth %>% filter(ID %in% test_ids)
  )
})
```

```{r inspect fold}
f <- folds[[1]]

train <- f$train
test  <- f$test

# check IDs don't overlap
intersect(unique(train$ID), unique(test$ID))

# Show first IDs
head(unique(train$ID), 10)
head(unique(test$ID), 10)

# show units inside one enterprise
train %>%
  filter(ID == train$ID[1]) %>%
  select(ID, unit) %>%
  distinct()

# compare structures
train %>%
  count(ID) %>%
  head()

test %>%
  count(ID) %>%
  head()

# number of units per entreprise
train %>%
  group_by(ID) %>%
  summarise(n_units = n()) %>%
  summary()
```

```{r use as usual}
errs <- sapply(folds, function(f) {
  
  train <- f$train
  test  <- f$test
  
  pred <- mean(train$QF185001)
  
  mean((test$QF185001 - pred)^2)
})

mean(errs)
```

```{r read file cv2}
data_cv2 <- read_parquet("data/data_synth_cv2.parquet")

dplyr::glimpse(data_cv2, width = 60)
```

### Time series / temporal data

Goal: Ensure realistic evaluation when predicting future values from past data

Problem: training data may include future observations which leads to data leakage and results in performance being overly optimistic

Solutions: - TimeSeriesSplit - train on past, test on future, - Rolling window - fixed-size moving training window, - Expanding window - training set grows over time

Key Message: If you train on the future to predict the past, you are no longer solving a real-world problem.

Example: the dataset has the same enterprise in several years

#### TimeSeriesSplit

This is a manual implementation of TimeSeriesSplit using an expanding window.

```{r Order data by time}

library(dplyr)

data_ts <- data_cv2 %>%
  arrange(ANO)

data_ts %>% count(ANO)
```

```{r Expanding window}

years <- sort(unique(data_ts$ANO))

# each fold trains on past years and tests on the next year
folds <- lapply(2:length(years), function(i) {
  
  train_years <- years[1:(i-1)]
  test_year   <- years[i]
  
  list(
    train = data_ts %>% filter(ANO %in% train_years),
    test  = data_ts %>% filter(ANO == test_year)
  )
})
```

```{r Inspect the expanding Fold}
f <- folds[[1]]

train <- f$train
test  <- f$test

range(train$ANO)
range(test$ANO)

# check no leakage
max(train$ANO) < min(test$ANO)
```

```{r Visualize train and test}
library(ggplot2)

plot_data <- bind_rows(
  train %>% mutate(set = "train"),
  test  %>% mutate(set = "test")
)

ggplot(plot_data, aes(x = ANO, fill = set)) +
  geom_bar(position = "dodge")
```

```{r Use as usual but with the expanding window}
errs <- sapply(folds, function(f) {
  
  train <- f$train
  test  <- f$test
  
  pred <- mean(train$QF185001)
  
  mean((test$QF185001 - pred)^2)
})

mean(errs)
```

#### Rolling Window

Evaluate models using recent past only, mimicking scenarios where older data becomes less relevant

```{r rolling folds}
library(dplyr)

data_ts <- data_cv2 %>%
  arrange(ANO)

years <- sort(unique(data_ts$ANO))
window <- 2  # number of past years to use

folds_roll <- lapply((window+1):length(years), function(i) {
  
  train_years <- years[(i-window):(i-1)]
  test_year   <- years[i]
  
  list(
    train = data_ts %>% filter(ANO %in% train_years),
    test  = data_ts %>% filter(ANO == test_year)
  )
})
```

```{r show multiple folds together}
library(ggplot2)

plot_folds <- lapply(seq_along(folds_roll), function(i) {
  
  f <- folds_roll[[i]]
  
  bind_rows(
    f$train %>% distinct(ANO) %>% mutate(set = "train", fold = i),
    f$test  %>% distinct(ANO) %>% mutate(set = "test",  fold = i)
  )
}) %>% bind_rows()

ggplot(plot_folds, aes(x = ANO, y = factor(fold), fill = set)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c(train = "steelblue", test = "orange")) +
  labs(y = "Fold", x = "Year")
```

```{r inspect one fold}

f <- folds_roll[[1]]

range(f$train$ANO)
range(f$test$ANO)

max(f$train$ANO) < min(f$test$ANO)
```

```{r Use as usual}

errs <- sapply(folds_roll, function(f) {
  
  train <- f$train
  test  <- f$test
  
  pred <- mean(train$QF185001)
  
  mean((test$QF185001 - pred)^2)
})

mean(errs)
```

#### Expanding windows

Assume that all past data remains relevant

```{r expanding folds}

library(dplyr)

data_ts <- data_cv2 %>%
  arrange(ANO)

years <- sort(unique(data_ts$ANO))

folds_exp <- lapply(2:length(years), function(i) {
  
  train_years <- years[1:(i-1)]
  test_year   <- years[i]
  
  list(
    train = data_ts %>% filter(ANO %in% train_years),
    test  = data_ts %>% filter(ANO == test_year)
  )
})

```

```{r}

library(ggplot2)

plot_folds_exp <- lapply(seq_along(folds_exp), function(i) {
  
  f <- folds_exp[[i]]
  
  bind_rows(
    f$train %>% distinct(ANO) %>% mutate(set = "train", fold = i),
    f$test  %>% distinct(ANO) %>% mutate(set = "test",  fold = i)
  )
}) %>% bind_rows()

ggplot(plot_folds_exp, aes(x = ANO, y = factor(fold), fill = set)) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c(train = "steelblue", test = "orange")) +
  labs(y = "Fold", x = "Year")
```

### Group + Time - You may need to use a combination of methods

Goal: Ensure evaluation is realistic when data has multiple entities (groups) observed over time

Problem - If we don't control both we may have: - Group leakage - same enterprise in train and test - Time leakage - future data present in train and the model may learn patterns that are impossible in the real world

Solution: Combine Split by time (respect time order) Grouping by entity (keeping the enterprises in group)

Key Message: When data has both group and time structure, you must respect both. Otherwise, leakage is unavoidable.

```{r Order data}

data_ts <- data_cv2 %>%
  arrange(ANO)
```

Each enterprise appears in several years but never shows in the future within the train

```{r create folds}

years <- sort(unique(data_ts$ANO))

# time first, groups preserved automatically
folds_gt <- lapply(2:length(years), function(i) {
  
  train_years <- years[1:(i-1)]
  test_year   <- years[i]
  
  list(
    train = data_ts %>% filter(ANO %in% train_years),
    test  = data_ts %>% filter(ANO == test_year)
  )
})
```

```{r}

library(ggplot2)
library(dplyr)

plot_folds_gt <- lapply(seq_along(folds_gt), function(i) {
  
  f <- folds_gt[[i]]
  
  bind_rows(
    f$train %>% mutate(set = "train", fold = i),
    f$test  %>% mutate(set = "test",  fold = i)
  ) %>%
    distinct(ID, ANO, set, fold)
}) %>% bind_rows()

ggplot(plot_folds_gt, aes(x = ANO, fill = set)) +
  geom_bar(position = "dodge") +
  facet_wrap(~ fold)
```

```{r Critical check}

f <- folds_gt[[1]]

# as expected the same enterprise is present in train and test
intersect(unique(f$train$ID), unique(f$test$ID))[1:50]

# but never in the same year
max(f$train$ANO) < min(f$test$ANO)
```

### Model selection vs performance estimation

Goal: Distinguish between choosing the best model and estimating its real-world performance

Problem: If you use the same data for both you optimize and evaluate on the same information which leads to optimistic bias and performance looks better than it really is

Solution: Separate the processes - Inner loop Model selection (tuning, comparing models) - Outer loop - Performance estimation (unbiased evaluation) This is typically done with nested cross-validation

Key Message: If you evaluate a model on the same data used to select it, you are measuring optimism, not performance.

```{r Performance estimation (Group CV)}

library(rsample)

folds <- group_vfold_cv(data_cv1, group = ID, v = 5)

outer_folds <- folds  
```

Nested Cross-Validation prevents overfitting during hyperparameter tuning

This looks like standard CV, but it is incorrect for this dataset.

Goal: Distinguish between choosing the best model and estimating its real-world performance

```{r Model selection, eval=FALSE}

```

### Spatial data

Examples: geolocation, environmental data

Problem: spatial autocorrelation

Solutions: Spatial blocking, Leave-location-out (nearby points ≠ independent)


