## Baseline: standard methods

Let's assume: - the enterprise only has one unit responding - the data is only from one year - data is independent and identically distributed

### Hold-out

Goal: Develop and evaluate a model quickly

Problem: A single split reduces available training data and produces a high-variance estimate of performance

Solution: Simple train/test split (e.g. 80/20). Fast and easy, but sensitive to how the data is split

Key message: A single split is a snapshot, not a guarantee. Model performance can change significantly depending on which observations end up in train vs test

```{r hold-out split}
library(rsample)
library(dplyr)
library(arrow)
data_cv0 <-  read_parquet("data/data_synth_cv0.parquet", as_data_frame = TRUE)
# set seed to ensure reproducibility
set.seed(123)

# do the split 80/20
split <- initial_split(data_cv0, prop = 0.8)

# use the split, to produce the train and test datasets
train_data <- training(split)
test_data  <- testing(split)
```

### K-Fold Cross-Validation

Goal: Obtain a more reliable estimate of model performance

Problem: A single train/test split is unstable and wastes data for training

Solution: Split data into K folds, train on K−1 folds and test on the remaining fold; repeat K times and average results. Good default when data is independent and identically distributed and there is no special structure.

Key message: K-fold CV reduces dependence on a single split and gives a more stable and data-efficient estimate of performance Split into K folds train on K-1, test on 1.

```{r k-folds}
set.seed(123)

folds <- vfold_cv(data_cv0, v = 5)

# keep only first fold
folds_1 <- folds$splits[[1]]

folds_1
```

```{r analyse fold}
fold_1 <- folds$splits[[1]]
fold_1

train_fold <- analysis(fold_1)
test_fold  <- assessment(fold_1)

length(folds$splits)
```
