
## Cross-validation based on data structure

### Imbalanced data

Goal: Ensure reliable evaluation of model performance when predicting an imbalanced target

Problem: In imbalanced datasets, random folds may not contain enough observations of rare target values, leading to biased or misleading performance estimates

Solution: Stratified K-Fold Cross-Validation, which preserves the distribution of the target variable (QF185001) in each fold

Key message: If you don't preserve the target distribution, you are no longer evaluating the same problem in each fold

```{r Stratified K-Fold}
library(rsample)

set.seed(123)

folds <- vfold_cv(data_cv0, v = 5, strata = QF185001)

split1 <- folds$splits[[1]]

train_data <- analysis(split1)
test_data  <- assessment(split1)
```

in the case where the variable is continuous like in our case we can use bins

```{r Stratified K-Fold with Bins}

# create the binned varaiable
data_cv0 <- data_cv0 |>
  mutate(QF185001_bin = ntile(QF185001, 5))

# lighter inspection
dplyr::glimpse(data_cv0, width = 60)

folds <- vfold_cv(data_cv0, v = 5, strata = QF185001_bin)
```

```{r examine bins}
data_cv0 %>%
  group_by(QF185001_bin) %>%
  summarise(
    min = min(QF185001, na.rm = TRUE),
    max = max(QF185001, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(QF185001_bin)
```

```{r visualize distribution}
library(ggplot2)

ggplot(data_cv0, aes(x = factor(QF185001_bin), y = QF185001 + 1)) +
  geom_boxplot() +
  scale_y_log10()
```

### Small datasets

Goal: Ensure reliable estimation of model performance when data is limited

Problem: With small datasets, model evaluation has high variance because each split contains very few observations, making results unstable and sensitive to data partitioning

Solution: - Leave-One-Out Cross-Validation (LOOCV): maximizes training data usage - Repeated K-Fold Cross-Validation: reduces variance by averaging across multiple random splits

Key Message: When data is scarce, the main risk is not bias but high variance in evaluation. Use resampling strategies that stabilize results

#### Leave-One-Out Cross-Validation (LOOCV)

uses all data efficiently but does not scale well, as it requires training one model per observation.

```{r LOOCV Step 1 - What are the folds?}

library(rsample)

# keeping only numeric predictors in the example 
data_num <- data_cv0 %>% select(where(is.numeric))

# making a small sample because we will have as many
# folds as observations
data_small <- data_num %>% slice_sample(n = 100) 

folds <- loo_cv(data_small)

length(folds$splits)

```

```{r LOOCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

nrow(train)
nrow(test)
```

```{r LOOCV Step 3 - How do we compute error?}

# errs stores the error for each fold
# since each fold has 1 test observation, each value is
# the prediction error for that observation
errs <- sapply(folds$splits, function(s) {
  train <- analysis(s)
  test  <- assessment(s)
  
  # lm is simple example model (linear regression)
  # not required for LOOCV
  # may replace it with any model (tree, random forest, etc.)
  m <- lm(QF185001 ~ ., data = train)
  pred <- predict(m, test)
  
  # mean squared error
  mean((test$QF185001 - pred)^2)
})
```

```{r}
# train many models (almost identical)
# each tested on one observation
# then average the errors
mean(errs)
```

#### Repeated K-Fold Cross-Validation

Repeated K-Fold reduces variance by averaging over multiple data splits.

```{r RKFCV Step 1 - What are the folds?}

library(rsample)

# split data into 5 folds
# repeat the whole process 10 times with different random splits
# Total models trained = 5 × 10 = 50
folds <- vfold_cv(data_cv0, v = 5, repeats = 10)

nrow(folds)

folds$splits[[1]]
```

```{r RKFCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

nrow(train)
nrow(test)
```

Train/test separation (\~80% / 20%)

```{r RKFCV Step 3 - How do we compute error?}

library(rpart)

# train many models on different splits
# compute error each time
# and average provide more stable estimate
errs <- sapply(folds$splits, function(s) {
  train <- analysis(s)
  test  <- assessment(s)

  train$ESTRATO <- factor(train$ESTRATO)
test$ESTRATO  <- factor(test$ESTRATO, levels = levels(train$ESTRATO))
  train$DD_CC_FF_COD <- factor(train$DD_CC_FF_COD)
  test$DD_CC_FF_COD  <- factor(test$DD_CC_FF_COD,
                               levels = levels(train$DD_CC_FF_COD))

  m <- rpart(QF185001 ~ ., data = train)
  pred <- predict(m, test)

  mean((test$QF185001 - pred)^2)
})

mean(errs)

```

```{r read file cv1}
data_cv1 <- read_parquet("data/data_synth_cv1.parquet")

dplyr::glimpse(data_cv1)
```

### Grouped / clustered data

Goal: Ensure that model evaluation does not leak information across related observations

Problem: Observations from the same entity (e.g. entreprise) are not independent If split randomly, the same entreprise can appear in both train and test, i.e. data leakage

Solution: Use Group K-Fold Cross-Validation, ensuring that all observations from the same group stay in the same fold

Key Message: If groups are split across folds, the model is evaluated on data it has effectively already seen

Examples: - multiple rows per user - entreprises, households, devices, patients

#### Group K-Fold Cross-Validation

Same entreprise appears multiple times, requires clustered data - ID - entreprise (group) - unit - establishment (multiple per entreprise)

```{r GKFCV Step 1 - What are the folds?}
library(rsample)

folds <- group_vfold_cv(data_cv1, group = ID, v = 5)
```

```{r GKFCV Step 2 - What is train vs test? }
s <- folds$splits[[1]]

train <- analysis(s)
test  <- assessment(s)

head(unique(train$ID), 10)
head(unique(test$ID), 10)

# check no overlap in groups
intersect(unique(train$ID), unique(test$ID))
```

For contrast we will compare with Random CV where we will see overlap

```{r Random CV}
folds_random <- vfold_cv(data_cv1, v = 5)

s <- folds_random$splits[[1]]

train_r <- analysis(s)
test_r  <- assessment(s)

intersect(unique(train_r$ID), unique(test_r$ID))[1:50]
```
